
---
title: "Literate Analysis with Ocean Water Samples"
subtitle: "coreR Session 05 — Section 5.7 Practice"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Introduction

This Quarto document follows the **Section 5.7 practice exercise** from coreR *Session 05 (Literate Analysis with Quarto)*: building a simple **prose + code** analysis using real seawater chemistry data (`BGchem2008data.csv`).\
Source: <https://learning.nceas.ucsb.edu/2024-10-coreR/session_05.html>

## About the data

The seawater chemistry dataset used here is `BGchem2008data.csv` (downloaded previously in the course).

-   Date downloaded: August 8, 2025
-   Download source: <https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A35ad7624-b159-4e29-a700-0c0770419941>

You will explore the nitrogen to phosphate ratio, which indicates the health of the ocean ecosystem. <https://beal-agulhas.earth.miami.edu/teaching/courses/lecture-five/index.html>

This analysis explores seawater chemistry data collected in 2008, focusing on the relationship between nitrogen and phosphate. In marine biology, the **Redfield Ratio** describes the consistent stoichiometric ratio of carbon, nitrogen, and phosphorus ($106:16:1$) found in marine phytoplankton and throughout the deep oceans. By calculating the ratio of Total Nitrogen to Phosphorus in this dataset, we can determine if the ecosystem is Nitrogen-limited or Phosphorus-limited. That may help us to hypothesize whether the marin species who are dependent on either phosphorous or nitrogen are in threat due to this imbalance composition. That may make us think why are these imbalance is created which is maybe due to pollutant, oil disposal domestic and industrial waste .

## Setup

Best practice: put all `library()` calls in one chunk near the top so collaborators can see what packages are required.

```{r}
library(here)
library(readr)
```

# Read in the data

Quarto runs R in the same way as a script: objects only exist after the chunks that create them have been run.

Using `here()` is a convenient way to build paths relative to your project root (e.g., your `.qmd` might live in `scripts/` while data lives in `data/`).

```{r}
bg_chem <- read_csv(here::here("/Users/muhammadsiraj/Documents/import-template-binsiraj19-blip/dat 806/Data/Raw", "BGchem2008data.csv"))

```

(Optional) Quick peek at the data:

```{r}
# Uncomment if you want more inspection output
 names(bg_chem)
# str(bg_chem)
 head(bg_chem)
#summary(bg_chem)
```

# Analysis
